---
title: "Class 09: Candy Mini-Project"
author: "Alyssa Duran (PID: A18550696)"
format: gfm
---

## In this mini-project, you will explore FiveThirtyEight's Halloween Candy dataset.

We will use lots of **ggplot** some basic stats, correlation analysis and PCA to make sense of the landscape of US candy - something hopefully more relatable than the proteomics and transcriptomics work that we will use these methods on throughout the rest of the course.


## Importing Candy Data

Our dataset sia CSV file so we use `read.csv()`.

```{r}
url <- "https://raw.githubusercontent.com/fivethirtyeight/data/master/candy-power-ranking/candy-data.csv"

candy_file <- read.csv(url)

candy <- read.csv(url, row.names=1)
head(candy)
```

> Q1. How many different candy types are in this dataset?

There are `r nrow(candy)` rows in this dataset.

```{r}
nrow(candy)
```

> Q2. How many fruity candy types are in the dataset?

There are `sum(candy$fruity)` fruity candy types in this dataset.

```{r}
sum(candy$fruity)
```


## What is Your Favorite Candy?

We can find the `winpercent` value for Twix by using its name to access the corresponding row of the dataset. For example the code for Twix is:

```{r}
candy["Twix", ]$winpercent

# An alternate way to do this is with the help of the dplyr package:
library(dplyr)

candy |> 
  filter(row.names(candy)=="Twix") |> 
  select(winpercent)
```

> Q3. What is your favorite candy (other than Twix) in the dataset and what is it’s winpercent value?

```{r}
candy |> 
  filter(row.names(candy)=="Skittles original") |>
  select(winpercent)
```

My favorite candy is skittles original and its `winpercent` value is `candy["Skittles original", ]$winpercent`.

> Q4. What is the winpercent value for “Kit Kat”?

```{r}
candy |> 
  filter(row.names(candy)=="Kit Kat") |>
  select(winpercent)
```

The `winpercent` value for Tootsie Roll Snack Bars is `candy["Kit Kat", ]$winpercent`.

> Q5. What is the winpercent value for “Tootsie Roll Snack Bars”?

```{r}
candy |> 
  filter(row.names(candy)=="Tootsie Roll Snack Bars") |> 
  select(winpercent)
```

The `winpercent` value for Tootsie Roll Snack Bars is `candy["Tootsie Roll Snack Bars", ]$winpercent`.

> Q6. Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?

Yes.

> Q7. What do you think a zero and one represent for the `candy$chocolate` column?

```{r}
library("skimr")
skim(candy)
```


## Exploratory Analysis

> Q8. Plot a histogram of winpercent values

```{r}
hist(candy$winpercent)

library(ggplot2)
ggplot(candy) + 
  aes(winpercent) + 
  geom_histogram(binwidth = 7, color = "salmon", fill = "pink") +
  theme_bw()
```

> Q9. Is the distribution of winpercent values symmetrical?

No.

> Q10. Is the center of the distribution above or below 50%?

```{r}
summary(candy$winpercent)
```

> Q11. On average is chocolate candy higher or lower ranked than fruit candy?

```{r}
mean(candy$winpercent[as.logical(candy$chocolate)])
mean(candy$winpercent[as.logical(candy$fruity)])
```

On average, chocolate candy is higher ranked than fruit candy.

> Q12. Is this difference statistically significant?

```{r}
t.test(candy$winpercent[as.logical(candy$chocolate)],
       candy$winpercent[as.logical(candy$fruity)])
```

The difference between chocolate and fruit candy is statistically significant.


## Overall Candy Rankings

Let’s use the base R `order()` function together with `head()` to sort the whole dataset by `winpercent`.

> Q13. What are the five least liked candy types in this set?

```{r}
ord.ind <- order(candy$winpercent)
head(candy[ord.ind,], n = 5)
```

> Q14. What are the top 5 all time favorite candy types out of this set?

```{r}
tail(candy[ord.ind,], n = 5)
```

To examine more of the dataset in this vain we can make a barplot to visualize the overall rankings. 

> Q15. Make a first barplot of candy ranking based on `winpercent` values.

```{r}
library(ggplot2)

ggplot(candy) + 
  aes(winpercent, rownames(candy)) +
  geom_col() 
```

> Q16. This is quite ugly, use the reorder() function to get the bars sorted by winpercent?

```{r}
ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy),winpercent)) +
  geom_col() +
  theme(axis.text.y = element_text(size = 3, hjust = 1))
```


## Time to Add Useful Color

We need a custom color vector.

```{r}
my_cols=rep("black", nrow(candy))
my_cols[as.logical(candy$chocolate==1)] = "chocolate"
my_cols[as.logical(candy$bar==1)] = "brown"
my_cols[as.logical(candy$fruity==1)] = "salmon"

ggplot(candy) + 
  aes(winpercent, 
      reorder(rownames(candy),winpercent)) +
  geom_col(fill=my_cols) +
  theme(axis.text.y = element_text(size = 3, hjust = 1))
```

> Q17. What is the worst ranked chocolate candy?

The worst ranked chocolate candy is Nik L Nip.

> Q18. What is the best ranked fruity candy?

The best ranked fruit candy is Reese's Peanut Butter cup.


## Taking a Look at pricepercent

What about value for money? What is the best candy for the least money? One way to get at this would be to make a plot of `winpercent` vs the `pricepercent` variable. The `pricepercent` variable records the percentile rank of the candy’s price against all the other candies in the dataset. Lower values are less expensive and higher values are more expensive.

```{r}
library(ggrepel)

# Plotting of win vs price
ggplot(candy) +
  aes(winpercent, pricepercent, label=rownames(candy)) +
  geom_point(col=my_cols) + 
  geom_text_repel(col=my_cols, size=3, max.overlaps = 3)
```

> Q19. Which candy type is the highest ranked in terms of winpercent for the least money - i.e. offers the most bang for your buck?

```{r}
ord_T <- order(candy$pricepercent, decreasing = TRUE)
head( candy[ord_T,c(11,12)], n=5 )
```

Hershey's Krackel is highest ranked in terms of `winpercent` of 62.28448, with the least money.

> Q20. What are the top 5 most expensive candy types in the dataset and of these which is the least popular?

```{r}
ord_F <- order(candy$pricepercent, decreasing = FALSE)
head( candy[ord_F,c(11,12)], n=5 )
```

Of the top 5 most expensive candy types in the dataset, Pixie Sticks are the least popular with a `winpercent` of 37.72234.


## Exploring the Correlation Structure

Now that we’ve explored the dataset a little, we’ll see how the variables interact with one another. We’ll use correlation and view the results with the **corrplot** package to plot a correlation matrix.

```{r}
library(corrplot)

cij <- cor(candy)
corrplot(cij)
```

> Q22. Examining this plot what two variables are anti-correlated (i.e. have minus values)?

`chocolate` and `fruit` variables are anti-correlated.

> Q23. Similarly, what two variables are most positively correlated?

`chocolate` and `bar` variables are the most positively correlated.


## Principal Component Analysis

Let’s apply PCA using the `prcomp()` function to our candy dataset remembering to set the `scale=TRUE` argument.

```{r}
pca <- prcomp(candy, scale = TRUE)
summary(pca)

plot(pca$x[,1:2], col=my_cols, pch=16)
```

We can make a much nicer plot with the ggplot2 package but it is important to note that ggplot works best when you supply an input data.frame that includes a separate column for each of the aesthetics you would like displayed in your final plot. To accomplish this we make a new data.frame here:

```{r}
# Making a new data-frame with our PCA results and candy data
my_data <- cbind(candy, pca$x[,1:3])

p <- ggplot(my_data) + 
        aes(x=PC1, y=PC2, 
            size=winpercent/100,  
            text=rownames(my_data),
            label=rownames(my_data)) +
        geom_point(col=my_cols)
p
```

Again we can use the **ggrepel** package and the function `ggrepel::geom_text_repel()` to label up the plot with non overlapping candy names like. We will also add a title and subtitle like so:

```{r}
library(ggrepel)

p + geom_text_repel(size=3, col=my_cols, max.overlaps = 3)  + 
  theme(legend.position = "none") +
  labs(title="Halloween Candy PCA Space",
       subtitle="Colored by type: chocolate bar (dark brown), chocolate other (light brown), fruity (red), other (black)",
       caption="Data from 538")
```

If you want to see more candy labels you can change the max.overlaps value to allow more overlapping labels or pass the ggplot object p to plotly like so to generate an interactive plot that you can mouse over to see labels: `library(plotly)` and `ggplotly(p)` # Does not work for pdf documents


Let’s finish by taking a quick look at PCA our loadings. Do these make sense to you? Notice the opposite effects of `chocolate` and `fruity` and the similar effects of `chocolate` and `bar` (i.e. we already know they are correlated).

```{r}
ggplot(pca$rotation) +
  aes(PC1,
      reorder(rownames(pca$rotation), PC1)) +
  geom_col()
```

> Q24. Complete the code to generate the loadings plot above. What original variables are picked up strongly by PC1 in the positive direction? Do these make sense to you? Where did you see this relationship highlighted previously?

The original variables including `fruity`, `pluribus`, and `hard` are strongly picked up by PC1 in the positive direction. This makes sense because PC1 is separating fruity, hard, “many-piece” candies from chocolate-bar style candies. We saw this earlier in the correlation matrix, where `chocolate` and `fruity` were strongly anti-correlated, and also in the `winpercent` comparison showing chocolate candies ranked much higher on average.


## Summary

> Q25. Based on your exploratory analysis, correlation findings, and PCA results, what combination of characteristics appears to make a “winning” candy? How do these different analyses (visualization, correlation, PCA) support or complement each other in reaching this conclusion?

A “winning” candy is most strongly associated with being chocolate-based, often a bar, and commonly including richer traits like caramel and peanut/almond, while being less fruity and less hard. The bar chart and top-5 list visually show that classic chocolate bars dominate the highest `winpercent` values, the correlation plot confirms chocolate clusters with bar, and PCA summarizes this same structure by placing chocolate/bar candies on one side of PC1 and fruity/hard candies on the other.




