---
title: "Class 12: Introduction to Genome Informatics Lab"
author: "Alyssa Duran (PID: A18005696)"
format: pdf
---

## Section 1: Proportion of G/G in a Population

Download a CSV file from Ensemble < https://useast.ensembl.org/Homo_sapiens/Variation/Sample?db=core;r=17:39894595-39895595;v=rs8067378;vdb=variation;vf=959672880#373531_tablePanel >

Here we read this CSV file:

```{r}
mxl <- read.csv("373531-SampleGenotypes-Homo_sapiens_Variation_Sample_rs8067378.csv")
head(mxl)

table(mxl$Genotype..forward.strand.) / nrow(mxl) *100
```


## Question 13 and 14:

Ready in file:

```{r}
expr <- read.table("rs8067378_ENSG00000172057.6.txt")
head(expr)

nrow(expr)
```

Make a boxplot:

```{r}
library(ggplot2)
ggplot(expr) + aes(geno, exp, fill=geno) + geom_boxplot(notch = T)
```

> Q14: Generate a boxplot with a box per genotype, what could you infer from the relative
expression value between A/A and G/G displayed in this plot? Does the SNP effect the
expression of ORMDL3? Hint: An example boxplot is provided overleaf – yours does not need
to be as polished as this one.

Individuals with the G/G genotype have significantly lower expression compared to those with the A/A genotype. This SNP is associated with reduced expression of the ORMDL3 gene. 


