---
title: "Class 5: Data Viz with ggplot"
author: "Alyssa Duran (PID: A18550696)"
format: gfm
toc: true
---

## Background

There are lot's of ways to make figures in R. These include so-called "base R" graphics (e.g. `plot()`) and tones of add-on packages like **ggplot2**.

For example here we make the same plot with both:

```{r}
head(cars, 3)
```

```{r}
plot(cars)
```
First I need to install the package with command `install.packages()`


> **N.B.** We never run an install cmd in a quarto code chunk or we will end up re-installing pakcages many many times - which is not what we want!

Every time you want to use one of these "add-on" packages we need to load it up in R with the `library()` function:

```{r}
library(ggplot2)
```

```{r}
ggplot(cars)
```

Every ggplot needs at least 3 things:

- The **data**, the stuff you want plotted
- The **aes**thitics, how the data map to the plot
- The **geom**etry,the type of plot

```{r}
ggplot(data = cars) +
  aes(x = speed, y = dist) +
  geom_point()
```

Add a line to better show relationship between speed and dist

```{r}
p <- ggplot(cars) +
  aes(x = speed, y = dist) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "Stopping Distance of Old Cars", 
       subtitle = "Data from the 'cars' object", 
       x = "Speed (mph)", 
       y = "Distance (ft)")
```

render it out
```{r}
p
```

```{r}
p + theme_bw()
```

## Gene Expression Plot

We can read the input data from the class website

```{r}
url <- "https://bioboot.github.io/bimm143_S20/class-material/up_down_expression.txt"
genes <- read.delim(url)
head(genes, 3)
```

A first version

```{r}
ggplot(genes) +
  aes(Condition1, Condition2) +
  geom_point()
```

```{r}
table(genes$State)
```

Second version colored by `State` so we can see the "up" and "down" significant genes compared to all the unchanging genes

```{r}
ggplot(genes) +
  aes(Condition1, Condition2, col = State) +
  geom_point()
```
Third version modifies the default colors to something more preferable

```{r}
ggplot(genes) +
  aes(Condition1, Condition2, col = State) +
  geom_point() +
  scale_colour_manual( values=c("blue",
                                "gray",
                                "red") ) +
  labs(x = "Control (no drug)",
       y = "Drug Treatment",
       title = "Gene Expression Changes upon GLP-1 Drug") +
  theme_bw()
```

## Going Further

Let's have a look at the famous **gapminder** dataset

```{r}
# File location online
url <- "https://raw.githubusercontent.com/jennybc/gapminder/master/inst/extdata/gapminder.tsv"

gapminder <- read.delim(url)
```

```{r}
head(gapminder, 3)
```

```{r}
ggplot(gapminder) +
  aes(gdpPercap, lifeExp, col = continent) +
  geom_point(alpha = 0.3)
```

Let's "facet" (i.e. make a separate plot) by continent rather than the big hot mess above

```{r}
ggplot(gapminder) +
  aes(gdpPercap, lifeExp, col = continent) +
  geom_point(alpha = 0.3) +
  facet_wrap(~continent)
```

## Custom Plots

How big is the gapminder dataset?

```{r}
nrow(gapminder)
```

I want to "filter" down to a subset of this data. I will use the **dplyr** package to help me.

First I need to install it and then load it up...
`install.packages("dplyr")` and then `library(dplyr)`

```{r}
library(dplyr)

gapminder_2007 <- filter(gapminder, year==2007)
head(gapminder_2007, 3)
```

```{r}
filter(gapminder_2007, country == "Ireland")
filter(gapminder_2007, year == 2007, country == "Ireland")
```

```{r}
filter(gapminder_2007, country == "United States")
```

> Q. Make a plot comparing 1977 and 2007 for all countries.

```{r}
input <- filter(gapminder, year %in% c(1977,2007))

ggplot(input) +
  aes(gdpPercap, lifeExp, col = continent) +
  geom_point() +
  facet_wrap(~year)
```




