---
title: "18: Pertussis and the CMI-PB project"
author: "Alyssa Duran (PID: A18550696)"
format: gfm
---

## 1. Investigating Pertussis Cases by Year

The United States Centers for Disease Control and Prevention (CDC) has been compiling reported pertussis case numbers since 1922 in their National Notifiable Diseases Surveillance System (NNDSS). We can view this data on the CDC website here: <https://www.cdc.gov/pertussis/surv-reporting/cases-by-year.html>

> Q1. With the help of the R “addin” package datapasta assign the CDC pertussis case number data to a data frame called cdc and use ggplot to make a plot of cases numbers over time.

```{r}
library(datapasta)
library(ggplot2)

cdc <- data.frame(Year = c(1922L,
1923L,1924L,1925L,1926L,1927L,1928L,
1929L,1930L,1931L,1932L,1933L,1934L,1935L,
1936L,1937L,1938L,1939L,1940L,1941L,
1942L,1943L,1944L,1945L,1946L,1947L,1948L,
1949L,1950L,1951L,1952L,1953L,1954L,
1955L,1956L,1957L,1958L,1959L,1960L,
1961L,1962L,1963L,1964L,1965L,1966L,1967L,1968L,1969L,1970L,1971L,1972L,1973L,
1974L,1975L,1976L,1977L,1978L,1979L,1980L,
1981L,1982L,1983L,1984L,1985L,1986L,
1987L,1988L,1989L,1990L,1991L,1992L,1993L,
1994L,1995L,1996L,1997L,1998L,1999L,
2000L,2001L,2002L,2003L,2004L,2005L,
2006L,2007L,2008L,2009L,2010L,2011L,2012L,
2013L,2014L,2015L,2016L,2017L,2018L,
2019L,2020L,2021L,2022L,2023L),
No..Reported.Pertussis.Cases = c(107473,164191,165418,152003,202210,181411,
161799,197371,166914,172559,215343,179135,
265269,180518,147237,214652,227319,103188,
183866,222202,191383,191890,109873,
133792,109860,156517,74715,69479,120718,
68687,45030,37129,60886,62786,31732,28295,
32148,40005,14809,11468,17749,17135,
13005,6799,7717,9718,4810,3285,4249,
3036,3287,1759,2402,1738,1010,2177,2063,
1623,1730,1248,1895,2463,2276,3589,
4195,2823,3450,4157,4570,2719,4083,6586,
4617,5137,7796,6564,7405,7298,7867,
7580,9771,11647,25827,25616,15632,10454,
13278,16858,27550,18719,48277,28639,
32971,20762,17972,18975,15609,18617,6124,
2116,3044,7063))

p <- ggplot(cdc, aes(x = Year, y =`No..Reported.Pertussis.Cases`)) + 
  geom_line() +
  geom_point() +
  labs(title = "Number of Reported Pertussis Cases Over Time",
       x = "Year",
       y = "Number of Reported Pertussis Cases") +
  theme_minimal()
p
```


## 2. A Tale of Two Vaccines (wP & aP)

Two types of pertussis vaccines have been developed: whole-cell pertussis (wP) and acellular pertussis (aP). The first vaccines were composed of ‘whole cell’ (wP) inactivated bacteria. The latter aP vaccines use purified antigens of the bacteria. These aP vaccines were developed to have less side effects than the older wP vaccines and are now the only form administered in the United States.

> Q2. Using the ggplot geom_vline() function add lines to your previous plot for the 1946 introduction of the wP vaccine and the 1996 switch to aP vaccine (see example in the hint below). What do you notice?

```{r}
p +
  geom_vline(xintercept = 1946, linetype = "dashed", color = "blue") +
  geom_vline(xintercept = 1996, linetype = "dashed", color = "red") +
  annotate("text", x = 1946,
           y = max(cdc$No..Reported.Pertussis.Cases),
           label = "wP",
           angle = 90,
           color = "blue") +
  annotate("text", x = 1996,
           y = max(cdc$No..Reported.Pertussis.Cases),
           label = "aP",
           angle = 90,
           color = "red")
```

> Q3. Describe what happened after the introduction of the aP vaccine? Do you have a possible explanation for the observed trend?

After the acellular pertussis (aP) vaccine was introduced around 1996, the number of reported pertussis cases began to increase again. A possible explanation is that the aP vaccine provided weaker/shorter-lasting immunity compared to the earlier whole-cell pertussis (wP) vaccine. Therefore, as immunity from the aP vaccine wanes over time, more people become susceptible to infection. 

## 3. Exploring CMI-PB Data

**Why is this vaccine-preventable disease on the upswing?** To answer this question we need to investigate the mechanisms underlying waning protection against pertussis. This requires evaluation of pertussis-specific immune responses over time in wP and aP vaccinated individuals.

```{r}
library(jsonlite)
subject <- read_json("https://www.cmi-pb.org/api/subject", simplifyVector = TRUE) 
head(subject, 3)
```

> Q4. How many aP and wP infancy vaccinated subjects are in the dataset?

```{r}
table(subject$infancy_vac)
```

> Q5. How many Male and Female subjects/patients are in the dataset?

```{r}
table(subject$biological_sex)
```

> Q6. What is the breakdown of race and biological sex (e.g. number of Asian females, White males etc…)?

```{r}
table(subject$biological_sex, subject$race)
```


## Side-Note: Working with Dates

Two of the columns of subject contain dates in the Year-Month-Day format. Recall from our last mini-project that dates and times can be annoying to work with at the best of times. However, in R we have the excellent lubridate package, which can make life allot easier. Here is a quick example to get you started:

```{r}
library(lubridate)
today()
today() - ymd("2000-01-01")
time_length( today() - ymd("2000-01-01"),  "years")
```

> Q7. Using this approach determine (i) the average age of wP individuals, (ii) the average age of aP individuals; and (iii) are they significantly different?

```{r}
library(lubridate)
library(dplyr)

# Use today's date to calculate age
subject$age <- today() - ymd(subject$year_of_birth)

# Separate groups
ap <- subject %>% filter(infancy_vac == "aP")
wp <- subject %>% filter(infancy_vac == "wP")

# Average ages
avg_age_wp <- mean(time_length(wp$age, "years"), na.rm = TRUE)
avg_age_ap <- mean(time_length(ap$age, "years"), na.rm = TRUE)

print(paste("Average wP age:", avg_age_wp))
print(paste("Average aP age:", avg_age_ap))

# Statistical test
t_test_result <- t.test(
  time_length(wp$age, "years"),
  time_length(ap$age, "years"),
  var.equal = TRUE
)

print(t_test_result)
```

> Q8. Determine the age of all individuals at time of boost?

```{r}
age_at_boost <- ymd(subject$date_of_boost) - ymd(subject$year_of_birth)
age_at_boost_years <- time_length(age_at_boost, "year")
head(age_at_boost_years)
```

> Q9. With the help of a faceted boxplot or histogram (see below), do you think these two groups are significantly different?

```{r}
ggplot(subject) +
  aes(time_length(age, "year"),
      fill=as.factor(infancy_vac)) +
  geom_histogram(show.legend=FALSE) +
  facet_wrap(vars(infancy_vac), nrow=2) +
  xlab("Age in years")
```

The histogram above show that the two groups are significantly different. The two groups are centered at a different age, the spread is different, and the groups share no overlaps.


## Joining Multiple Tables

Read the specimen and ab_titer tables into R and store the data as specimen and titer named data frames.

```{r}
# Complete the API URLs:
specimen <- read_json("https://www.cmi-pb.org/api/specimen", simplifyVector = TRUE)
titer <- read_json("https://www.cmi-pb.org/api/plasma_ab_titer", simplifyVector = TRUE) 
```

> Q9. Complete the code to join specimen and subject tables to make a new merged data frame containing all specimen records along with their associated subject details:

```{r}
meta <- inner_join(specimen, subject)
dim(meta)
head(meta)
```

> Q10. Now using the same procedure join meta with titer data so we can further analyze this data in terms of time of visit aP/wP, male/female etc.

```{r}
abdata <- inner_join(titer, meta)
dim(abdata)
```

> Q11. How many specimens (i.e. entries in abdata) do we have for each isotype?

```{r}
table(abdata$isotype)
```

> Q12. What are the different $dataset values in abdata and what do you notice about the number of rows for the most “recent” dataset?

```{r}
table(abdata$dataset)
```

There are four different $dataset values in abdata. The most recent dataset has the least amount of rows. The decrease in rows may be due to incomplete data reporting or changes in data reporting.


## 4. Examine IgG Ab Titer Levels

Now using our joined/merged/linked abdata dataset filter() for IgG isotype.

```{r}
igg <- abdata %>% filter(isotype == "IgG")
head(igg)
```

> Q13. Complete the following code to make a summary boxplot of Ab titer levels (MFI) for all antigens:

```{r}
ggplot(igg) +
  aes(MFI_normalised, antigen) +
  geom_boxplot() +
  xlim(0,75) +
  facet_wrap(vars(visit), nrow=2)
```

> Q14. What antigens show differences in the level of IgG antibody titers recognizing them over time? Why these and not others?

The antigens that show differences in IgG antibody titers over time are FIM2/3 and FHA. The change in antigens suggests a recall response after vaccination or that its antibody levels for thees antigens increase or vary across visits. Antigens OVA and PT show little difference and more overlap because they do not trigger a strong recall response or they have consistent baseline antibody levels.

> Q15. Filter to pull out only two specific antigens for analysis and create a boxplot for each. You can chose any you like. Below I picked a “control” antigen (“OVA”, that is not in our vaccines) and a clear antigen of interest (“PT”, Pertussis Toxin, one of the key virulence factors produced by the bacterium B. pertussis).

```{r}
filter(igg, antigen=="OVA") %>%
  ggplot() +
  aes(MFI_normalised, col=infancy_vac) +
  geom_boxplot(show.legend = F) +
  facet_wrap(vars(visit)) +
  theme_bw()
```

> Q16. What do you notice about these two antigens time courses and the PT data in particular?

The two antigens show changes in antibody titers over time, with levels generally increasing after vaccination and then leveling off or declining at later visits. The PT data shows a clear increase in IgG levels, indicating a strong immune response to this antigen over time.

> Q17. Do you see any clear difference in aP vs. wP responses?

Yes, there is a difference between aP and wP responses. Individuals who received the aP vaccine tend to show higher or more consistent antibody responses to PT, whereas responses in wP individuals appear lower or more variable.

```{r}
abdata.21 <- abdata %>% filter(dataset == "2021_dataset")

abdata.21 %>% 
  filter(isotype == "IgG",  antigen == "PT") %>%
  ggplot() +
    aes(x=planned_day_relative_to_boost,
        y=MFI_normalised,
        col=infancy_vac,
        group=subject_id) +
    geom_point() +
    geom_line() +
    geom_vline(xintercept=0, linetype="dashed") +
    geom_vline(xintercept=14, linetype="dashed") +
  labs(title="2021 dataset IgG PT",
       subtitle = "Dashed lines indicate day 0 (pre-boost) and 14 (apparent peak levels)")
```

> Q18. Does this trend look similar for the 2020 dataset?

```{r}
abdata.20 <- abdata %>% 
  filter(dataset == "2020_dataset")

abdata.20 %>%
  filter(isotype == "IgG", antigen == "PT") %>%
  ggplot(aes(x = planned_day_relative_to_boost,
             y = MFI_normalised,
             color = infancy_vac,
             group = subject_id)) +
  geom_point() +
  geom_line() +
  geom_vline(xintercept = 0, linetype = "dashed") +
  geom_vline(xintercept = 14, linetype = "dashed") +
  labs(
    title = "2020 dataset IgG PT",
    subtitle = "Dashed lines indicate day 0 (pre-boost) and 14 (apparent peak levels)"
  ) +
  theme_bw()
```

By comparing the 2020 and 2021 datasets, it becomes apparent that the 2020 dataset's has a longer follow-up period. Additionally, PT antibody signals appear noticeably higher in the 2020 dataset for both the aP and wP groups.


## 5. Obtaining CMI-PB RNASeq Data

For RNA-Seq data the API query mechanism quickly hits the web browser interface limit for file size. We will present alternative download mechanisms for larger CMI-PB datasets in the next section. However, we can still do “targeted” RNA-Seq querys via the web accessible API.

```{r}
url <- "https://www.cmi-pb.org/api/v2/rnaseq?versioned_ensembl_gene_id=eq.ENSG00000211896.7"
rna <- read_json(url, simplifyVector = TRUE) 

ssrna <- inner_join(rna, meta)
```

> Q19. Make a plot of the time course of gene expression for IGHG1 gene (i.e. a plot of visit vs. tpm).

```{r}
ggplot(ssrna) +
  aes(visit, tpm, group=subject_id) +
  geom_point() +
  geom_line(alpha=0.2)
```

> Q20. What do you notice about the expression of this gene (i.e. when is it at it’s maximum level)?

The expression of this gene increases after vaccination and reaches its highest levels around the early to mid visits, particularly around visit 4, before decreasing again at later visits. Visit 8 also shows a peak, however it is most likely an outlier. This suggests the gene is most active shortly after the immune response is stimulated.

> Q21. Does this pattern in time match the trend of antibody titer data? If not, why not?

This pattern does not exactly match the antibody titer trend. Gene expression tends to peak earlier because it reflects the immediate activation of immune cells after vaccination, while antibody titers increase later as the immune system produces and accumulates antibodies over time.

We can dig deeper and color and/or facet by infancy_vac status:

```{r}
ggplot(ssrna) +
  aes(tpm, col=infancy_vac) +
  geom_boxplot() +
  facet_wrap(vars(visit))
```

There is however no obvious wP vs. aP differences here even if we focus in on a particular visit:

```{r}
ssrna %>%  
  filter(visit==4) %>% 
  ggplot() +
    aes(tpm, col=infancy_vac) + geom_density() + 
    geom_rug() 
```









