---
title: "Class 6: R Functions Homework"
author: "Alyssa Duran (PID: A18550696)"
format: pdf
---

## R Functions

> First, install packages by using `install.packages("bio3d)` and `library()`

```{r}
library(bio3d)
```

> Section 1A

```{r}
# (A. Can you improve this analysis code?
df <- data.frame(a=1:10, b=seq(200,400,length=10),c=11:20,d=NA)
df$a <- (df$a - min(df$a)) / (max(df$a) - min(df$a))
df$b <- (df$b - min(df$a)) / (max(df$b) - min(df$b))
df$c <- (df$c - min(df$c)) / (max(df$c) - min(df$c))
df$d <- (df$d - min(df$d)) / (max(df$a) - min(df$d)) 
```

> Section 1B

```{r}
# Can you improve this analysis code?
s1 <- read.pdb("4AKE") # kinase with drug
s2 <- read.pdb("1AKE") # kinase no drug
s3 <- read.pdb("1E4Y") # kinase with drug
s1.chainA <- trim.pdb(s1, chain="A", elety="CA")
s2.chainA <- trim.pdb(s2, chain="A", elety="CA")
s3.chainA <- trim.pdb(s1, chain="A", elety="CA")
s1.b <- s1.chainA$atom$b
s2.b <- s2.chainA$atom$b
s3.b <- s3.chainA$atom$b
plotb3(s1.b, sse=s1.chainA, typ="l", ylab="Bfactor")
plotb3(s2.b, sse=s2.chainA, typ="l", ylab="Bfactor")
plotb3(s3.b, sse=s3.chainA, typ="l", ylab="Bfactor")
```


```{r}
hc <- hclust( dist( rbind(s1.b, s2.b, s3.b) ) )
plot(hc)
```

## Q6. How would you generalize the original code above to work with any set of input protein structures?

> Inputs to the Function

- `pdb_ids`: Character vector of PDB IDs (e.g., `c("4AKE", "1AKE", "1E4Y")`)
- `chain`: Protein chain to analyze (default = `"A"`)
- `elety`: Atom type to extract (default = `"CA"` for alpha carbons)
- `plot_individual`: Whether to plot individual B-factor plots (default = `TRUE`)
- `plot_cluster`: Whether to plot hierarchical clustering dendrogram (default = `TRUE`)

```{r}
# Example call:
analyze_bfactor("4AKE")
analyze_bfactor("1AKE")
analyze_bfactor("1E4Y")
```

> What the Function Does and How to Use It

This function reads multiple protein PDB files, extracts B-factor values, stores them for all proteins, and plots B-factors for the specified protein. The function generalizes the original code to work with any set of proteins.

To use the function, first install packages by `install.packages("bio3d")` and `library()` Then, use the function with PDB IDs by `analyze_bfactor()`. Changing chain, atom type, and protein plots can be done as well.

```{r}
# Example: 
analyze_bfactor(c("4AKE", "1AKE", "1E4Y"), chain = "A")
```

> Output of the Function

The function returns a name list where each value is a vector of B-factor values where the name is the PDB IDs. The function displays a B-factor plot for the specified protein

- Each element name is a PDB ID
- Each element value is a numeric vector of B-factor values
- Individual B-factor plots are displayed (if `plot_individual = TRUE`)
- A hierarchical clustering dendrogram is displayed for multiple structures (if `plot_cluster = TRUE`)

```{r}
analyze_bfactor <- function(pdb_ids, chain = "A", elety = "CA", plot_individual = TRUE, plot_cluster = TRUE) {
  
  # Load bio3d package
  library(bio3d)
  
  # Initialize list to store B-factor values
  bfactor_list <- list()
  
  # Loop through each PDB ID and read PDB file
  for (pdb_id in pdb_ids) {
    cat("Processing:", pdb_id, "\n")
   
    pdb <- read.pdb(as.character(pdb_id)) 
    
    # Trim to specified chain and atom type
    pdb_trim <- trim.pdb(pdb, chain = chain, elety = elety)
    
    # Extract B-factor values
    bfac <- pdb_trim$atom$b
    
    # Store in list with PDB ID as name
    bfactor_list[[pdb_id]] <- bfac
    
    # Plot individual B-factor plot if requested
    if (plot_individual) {
      plotb3(
        bfac,
        sse = pdb_trim,
        typ = "l",
        ylab = "B-factor",
        main = paste("B-factor plot for", pdb_id)
      )
    }
  }
  
  # Perform hierarchical clustering if multiple structures and clustering is requested
  if (length(pdb_ids) > 1 && plot_cluster) {
    # Combine B-factors into matrix
    bfactor_matrix <- do.call(rbind, bfactor_list)
    
    # Perform hierarchical clustering
    hc <- hclust(dist(bfactor_matrix))
    
    # Plot dendrogram
    plot(hc, main = "Hierarchical Clustering of B-factors", 
         xlab = "Protein Structures", ylab = "Distance")
  }
  
  # Return named list of B-factors
  return(bfactor_list)
}
results_single <- analyze_bfactor("4AKE")

results_multiple <- analyze_bfactor(c("4AKE", "1AKE", "1E4Y"), chain = "A")

results_cluster_only <- analyze_bfactor(
  c("4AKE", "1AKE", "1E4Y"), 
  plot_individual = FALSE, 
  plot_cluster = TRUE
)

results <- analyze_bfactor(c("4AKE", "1AKE", "1E4Y"))
results$`4AKE`
```



